Learning \strips\ action models from {\em state-invariants}

### 2019-02-05 doc/ijcai-19-deductivelearning/deductivelearning.tex
first updates to section 3 ijcai19 paper on deductive learning

\begin{abstract}
\end{abstract}

@@

The specification of a planning action model is a complex process that limits, too often, the application of model-based planning to real-world tasks~\cite{kambhampati:modellite:AAAI2007}. The {\em machine learning} of action models can relieve this {\em knowledge acquisition bottleneck} of {\em AI planning} and nowadays, there exists a wide range of effective approaches for learning planning action models~\cite{arora:amodels:ker2018}. Many of the most successful approaches for learning planning action models are however purely {\em inductive}~\cite{yang2007learning,pasula2007learning,mourao2010learning,zhuo2013action}, meaning that their performance depends on the {\em amount} and {\em quality} of the input examples (namely observations of plan executions generated with the aimed action model).

This paper proposes the exploitation of {\em deductive} knowledge in the form of {\em state-invariants} (i.e logic formulae that specify constraints about the possible states of a given domain) to reduce the negative effects of insufficient input observations when learning \strips\ action models. Given an action model state-of-the-art planners use to compute {\em state-invariants} to reduce their search space and make the planning process more efficient~\cite{helmert2009concise}. In this paper we follow the oposite direction and exploit {\em state-invariants} to learn a planning action model.

@@

\section{Learning action models from {\em state-invaraints}}
\label{sec:evaluation}

We define the task of learning action models as a tuple $\Lambda=\tup{M,\Phi,\mathcal{O}(\tau)}$, where:
\begin{itemize}
\item $M$ is the {\em space of possible action model}. This set is the {\em full} space of action models, when learning from scratch, or {\em partially specified}, when some fragments of the action models are known a priori. A set of action models can be defined {\em explicitly}, enumerating all the models that belong to the set or {\em implicitly}, enumerating all the constraints that must satisfy any model that belongs to the set. A {\em partially specified} \strips\ action model is then a formalism for the {\em implicit} representation of a set of \strips\ schemes~\cite{sreedharan2018handling}. 
\item $\Phi$, a set of {\em state-invariants} that constrain the set of possible states in the given domain.
\item $\mathcal{O}(\tau)$ is an observation of a trajectory $\tau(\pi,P)$ produced by the execution of an unknown plan $\pi$ that solves $P_\mathcal{O}$ (i.e. the classical planning problem $P_\mathcal{O}=\tup{F,A[\cdot],s_0^o,s_m^o}$ where $A[\cdot]$ is a set of actions s.t., the semantics of each action $a\in A[\cdot]$ is unknown since functions $\rho$ and/or $\theta$ of $a$ are undefined). 
\end{itemize}

A {\em solution} to a $\Lambda=\tup{M,\Phi,\mathcal{O}(\tau)}$ learning task is a model $\mathcal{M}'\in M$ that is {\em consistent} with the given input knowledge $\Phi$ and $\mathcal{O}(\tau)$. This means that let there exists a plan $\pi$ such that the observation $\mathcal{O}(\tau)$ is consistent with $\pi$ when the dynamics of $A[\cdot]$ is given by $\mathcal{M}'\in M$ and such that any state produced in the trajectory $\tau(\pi,P_\mathcal{O})$ satisfies the {\em state-invariants} $\Phi$.

%They compute an action model that maximizes some notion of {\em statistical consistency} over a set of observations of plan executions so output an action model that is {\em consistent} with the input knowledge.

Next we show that the set $M$ of possible action models can be encoded as a set of propositional variables and a set of constraints over those variables. Then, we show how to exploit this encoding to solve the $\Lambda=\tup{M,\Phi,\mathcal{O}(\tau)}$ learning task with an off-the-shelf classical planner.

@@

\subsection{Learning \strips\ action models with classical planning}
Given a $\Lambda=\tup{M,\Phi,\mathcal{O}(\tau)}$ where $\Phi$ is a set of {\em domain mutex} $\phi\in\Phi$, we create a classical planning problem $P'=\tup{F',A',I,G}$ such that:
\begin{itemize}
\item $F'$ extends $F$ with a fluent $mode_{insert}$, to indicate whether action models are being programmed, and the fluents for the propositional encoding of the corresponding space of STRIPS action models. This is a set of fluents of the type $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of FOL interpretations of predicates $\Psi$ over the corresponding action parameters $pars(\xi)$.

@@

\begin{small}
\begin{align*}
\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{p}(\xi)\implies p(\omega)\}_{\forall p\in\Psi_\xi},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{pre_{p}(\xi)\wedge eff_{p}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi_\xi},\\
&\{\neg pre_{p}(\xi)\wedge eff_{p}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi\xi}\},\\
&\{\emptyset\}\rhd\{\neg mode_{insert}\}.
\end{align*}
\end{small}

>>>>

\begin{abstract}
This paper addresses the learning of action models from {\em deductive} knowledge with the form of {\em state-invariants} (i.e logic formulae that specify constraints about the possible states of a given domain) to reduce the negative effects of insufficient input observations. Given an action model, state-of-the-art planners use it to infer {\em state-invariants} that reduce the planner search space and make the planning process more efficient~\cite{helmert2009concise}. In this paper we follow the opposite direction and exploit {\em state-invariants} to learn a planning action model.
\end{abstract}

@@

The specification of planning action models is a complex process that limits, too often, the application of {\em model-based planning} to real-world tasks~\cite{kambhampati:modellite:AAAI2007}. The {\em machine learning} of action models can relieve the {\em knowledge acquisition bottleneck} of planning and nowadays, there exists a wide range of effective approaches for learning action models~\cite{arora:amodels:ker2018}. Many of the most successful approaches for learning planning action models are however purely {\em inductive}~\cite{yang2007learning,pasula2007learning,mourao2010learning,zhuo2013action}, meaning that their performance depends on the {\em amount} and {\em quality} of the input examples (namely observations of plan executions generated by the aimed action model).

This paper addresses the learning of action models from {\em deductive} knowledge with the form of {\em state-invariants} (i.e logic formulae that specify constraints about the possible states of a given domain) to reduce the negative effects of insufficient input observations. Given an action model, state-of-the-art planners use it to infer {\em state-invariants} that reduce the planner search space and make the planning process more efficient~\cite{helmert2009concise}. In this paper we follow the opposite direction and exploit {\em state-invariants} to learn a planning action model.

@@

\section{Learning \strips\ action models from {\em state-invaraints}}
\label{sec:learning}

We define the task of learning an action model from {\em state-invaraints} as a tuple $\Lambda=\tup{P,\Phi,M}$, where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$, is a {\em classical planning problem} where $A[\cdot]$ is a set of actions s.t., the semantics of each action $a\in A[\cdot]$ is unknown (i.e. the functions $\rho$ and/or $\theta$ of $a$ are undefined).
\item $\Phi$, a set of {\em state-invariants} that constrain the set of possible states in the given domain.
\item $M$ is the {\em space of possible action model} for the $A[\cdot]$ actions. This set is the {\em full} space of action models, when learning from scratch, or a {\em partially specified action model}, when some fragments of the aimed action model are a priori known. A set of action models can be defined {\em explicitly}, enumerating all the models that belong to the set or {\em implicitly}, enumerating all the constraints that must satisfy any model that belongs to the set. A {\em partially specified} \strips\ action model is then a formalism for the {\em implicit} representation of a set of \strips\ schemes~\cite{sreedharan2018handling}.
\end{itemize}

A {\em solution} to a $\Lambda=\tup{P,\Phi,M}$ learning task is a model $\mathcal{M}'\in M$ such that there exists a plan $\pi$ satisfying that any state traversed in the trajectory $\tau(\pi,P)$ satisfies all the {\em state-invariants} in $\Phi$. Inductive approaches for the learning planning action models compute an action model that maximizes some notion of {\em statistical consistency} over a set of observations of plan executions so output an action model in our case a solution to the addressed learning task is an action model that is {\em consistent} with the input knowledge.

Next we show that the set $M$ of possible action models can be encoded as a set of propositional variables and a set of constraints over those variables. Then, we show how to exploit this encoding to solve a $\Lambda=\tup{P,\Phi,M}$ learning task with an off-the-shelf classical planner.

@@

\subsection{Learning \strips\ action models with classical planning}
Given a $\Lambda=\tup{P,\Phi,M}$ where $\Phi$ is a set of {\em domain mutex} $\phi\in\Phi$, we create a classical planning problem $P'=\tup{F',A',I,G}$ such that:
\begin{itemize}
\item $F'$ extends $F$ with a fluent ${\tt inconsistent}$, to indicate whether an action model is {\em inconsistent} with the input {\em state-invariants}, a fluent $mode_{insert}$, to indicate whether action models are being programmed, and the fluents for the propositional encoding of the corresponding space of STRIPS action models. This is a set of fluents of the type $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of FOL interpretations of predicates $\Psi$ over the corresponding action parameters $pars(\xi)$.

@@

\begin{small}
\begin{align*}
\pre(\mathsf{apply_{\xi,\omega}})=&\{\emptyset\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{pre_{p}(\xi)\wedge \neg p(\omega)\}_{\forall p\in\Psi_\xi},\\
&\{pre_{p}(\xi)\wedge eff_{p}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi_\xi},\\
&\{\neg pre_{p}(\xi)\wedge eff_{p}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi\xi}\},\\
&\{\emptyset\}\rhd\{\neg mode_{insert}\}.
\end{align*}
\end{small}

\section{Learning from {\em state-invariants} and observations of plan executions}
\label{sec:observations}
Most of the approachs for leanning planning action mdoels compute an action model that maximizes some notion of {\em statistical consistency} over a set of observations of plan executions so output an action model that is {\em consistent} with the input knowledge.

\subsection{Extending the compilation to deal with {\em plan recognition as planning}}
The work on {\em plan recognition as planning} usually assumes an observation model that is referred only to logs of executed actions. However, the approach applies also to more expressive observation models that consider state observations as well, like the observation model defined above, with a simple three-fold extension:
\begin{itemize}
\item One fluent $\{validated_j\}_{0\leq j\leq m}$ to point at every $s_j^o\in\mathcal{O}(\tau)$ state observation. Two fluents, $at_i$ and $next_{i,i+1}$, {\small $1\leq i \leq n$}, to iterate through the $n$ observed actions of $\tau$. The former is used to ensure that actions are executed in the same order as they are observed in $\tau$. The latter is used to iterate to the next planning step when solving $P_{\Lambda}$.
\item Adding $at_1$ and $\{next_{i,i+1}\}$, {\small $1\leq i \leq n$} to the initial state and $validated_m$ to every possible goal $G\in G[\cdot]$ to constrain solution plans $\pi^\top$ to be consistent with all the state observations.

\item When the input plan trace contains observed actions, the extra conditional effects $\{at_{i},plan(name(a_i),\Omega^{ar(a_i)},i)\}\rhd\{\neg at_{i},at_{i+1}\}_{\forall i\in [1,n]}$ are included in the $\mathsf{apply_{\xi,\omega}}$ actions to ensure that actions are applied in the same order as they appear in $\tau$.\\

\item Actions for {\em validating} the partially observed state $s_j\in\tau$, {\tt\small $1\leq j< m$}. These actions are also part of the postfix of the solution plan $\pi_\Lambda$ and they are aimed at checking that the observable data of the input plan trace $\tau$ follows after the execution of the apply actions.

\item One $\mathsf{validate_{j}}$ action to constraint $\pi^\top$ to be consistent with the $s_j^o\in\mathcal{O}(\tau)$ input state observation, {\small $(1\leq j\leq m)$}.  
\end{itemize}
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{j}})=&s_j^o\cup\{validated_{j-1}\},\\
\cond(\mathsf{validate_{j}})=&\{\emptyset\}\rhd\{\neg validated_{j-1}, validated_j\}.
\end{align*}
\end{small}
